### 2020-06-11 writeup/proof_main.tex
revise

For the isotropic model in Section \ref{sec_insight}, we actually have an easier and sharper bound than Theorem \ref{thm_model_shift} as follows. 
\begin{lemma}[The isotropic model]\label{prop_model_shift_tight}
		In the setting of Theorem \ref{thm_model_shift}, assume that $\Sigma_1 =\id$,
		$\beta_2$ is a random vector with i.i.d. entries with mean $0$, variance $\kappa^2$ and all moments, and $\beta_1$ is a random vector such that $(\beta_1 - \beta_2)$ is a random vector with i.i.d. entries with mean $0$, variance $d^2$ and all moments. Denote
		$\Delta^\star_{\bias} := \bigbrace{(1 - \hat{v})^2 \kappa^2 + d^2} \bigtr{\Pi(\hat v)}.$
	Then we have with high probability,
		\begin{align*}
			\te(\hat{\beta}_t^{\MTL}) \le \te(\hat{\beta}_t^{\STL})\quad \text{ when: }\ \  & \Delta_{\vari} \ge  (\al_+^2(\rho_1)+\oo(1))\cdot  \Delta^\star_{\bias}, \\
			\te(\hat{\beta}_t^{\MTL}) \ge \te(\hat{\beta}_t^{\STL})\quad \text{ when: }\ \  & \Delta_{\vari} \le (\al_-^2(\rho_1)-\oo(1))\cdot  \Delta^\star_{\bias}.
		\end{align*}
\end{lemma}

The proof of Theorem \ref{thm_model_shift} is based on Lemma \ref{lem_cov_shift}, Lemma \ref{lem_cov_derivative}, and the following bound on the singular values of $Z_1$: for any fixed $\e>0$, we have
\begin{align}
\al_-(\rho_1) - \OO(p^{-1/2+e})  \preceq {n_1^{-1}}{Z_1^T Z_1}  \preceq   \al_+(\rho_1) + \OO(p^{-1/2+e}) \quad \text{w.h.p.}  \label{eq_isometric}
\end{align}
%with high probability. 
In fact, $n_1^{-1}Z_1^TZ_1$ is a standard sample covariance matrix, and it is well-known that its eigenvalues are inside the support of the Marchenko-Pastur law $[\al_-(\rho_1)-\oo(1) ,\al_+(\rho_1)+\oo(1)]$ with probability $1-\oo(1)$ \cite{No_outside}.
The estimate \eqref{eq_isometric} provides tight bounds for the concentration error.
The result is formally stated in Lemma \ref{SxxSyy} in Section \ref{sec_maintools}.

@@

\begin{remark}
One can see from the above proof that the main error, $\delta$, of Theorem \ref{thm_model_shift} comes from approximating $\delta_{\bias}$ by $\wt\delta_{\bias}$ in \eqref{bounddelta-}. In order to improve this estimate and obtain an exact asymptotic result, one needs to study the singular value distribution of the following random matrix:
$$(X_1^{\top}X_1)^{-1}X_2^{\top}X_2 +  {v}^2 .$$
In fact, the eigenvalues of $\cal X:=(X_1^{\top}X_1)^{-1}X_2^{\top}X_2$ have been studied in the name of Fisher matrices; see e.g. \cite{Fmatrix}. However, since $\cal X$ is not symmetric, it is known that the singular values of $\cal X$ are different from its eigenvalues. To the best of our knowledge, the asymptotic singular value behavior of $\cal X$ is still unknown in random matrix theory literature, and the study of the singular values of $\cal X + v^2$ will be even harder. We leave this problem to future study.
\end{remark}

By replacing \eqref{bounddelta-} with a tighter bound in Step II of the above proof, we can conclude the proof of Lemma \ref{prop_model_shift_tight}.

\begin{proof}[Proof of Lemma \ref{prop_model_shift_tight}]
For any fixed $v\in \R$, $\beta_1 - v\beta_2$ is a random vector with i.i.d. entries with mean $0$ and variance $(1-v)^2\kappa^2 + d^2$. Then using the concentration result, Lemma \ref{largedeviation}, we get that for any constant $\e>0$,
\begin{align}
&\left|\delta_{\bias}(v)- [(1-v)^2\kappa^2 + d^2]\tr (\cal K^\top \cal K)\right| \nonumber\\
&=  \left|(\beta_1 - {v} \beta_2)^\top \cal K^\top \cal K (\beta_1 - {v} \beta_2) - [(1-v)^2\kappa^2 + d^2]\tr (\cal K^\top \cal K)\right| \nonumber\\
&\le p^\e [(1-v)^2\kappa^2 + d^2] \left\{\tr\left[(\cal K^\top \cal K)^2\right]\right\}^{1/2} \lesssim p^{1/2+\e} [(1-v)^2\kappa^2 + d^2],\label{anotherbeta}
\end{align}
where we denoted $\cal K:=v\Sigma_2^{1/2}({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1} X_1^{\top}X_1$, and in the last step we used $\|\cal K\|=\OO(1)$ by \eqref{eq_isometric}. Now for $\tr (\cal K^\top \cal K)$, we rewrite it as
\begin{align*}
v^2 \tr \left[ ({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\Sigma_2({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}(X_1^\top X_1)^2 \right].
\end{align*}
Bounding $(X_1^\top X_1)^2=(Z_1^\top Z_1)^2$ using \eqref{eq_isometric} again, we obtain that
\be\label{anotherwtbeta} \delta^\star_{\bias}(v) \cdot (\al_-^2(\rho_1)-\OO(p^{-1/2+\e})) \le  [(1-v)^2\kappa^2 + d^2]\tr (\cal K^\top \cal K) \le \delta^\star_{\bias}(v) \cdot  (\al_+^2(\rho_1)+\OO(p^{-1/2+\e})),\ee
where 
$$\delta^\star_{\bias}(v):=n_1^2 v^2 [(1-v)^2\kappa^2 + d^2]\tr \left[ ({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\Sigma_2({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\right].$$
Note that $\delta^\star_{\bias}(v) \sim p$, hence combining \eqref{anotherbeta} and \eqref{anotherwtbeta} we get 
\be\label{replaceest}
  \delta^\star_{\bias}(v) \cdot (\al_-^2(\rho_1)-\OO(p^{-1/2+\e})) \le  \delta_{\bias}(v)\le \delta^\star_{\bias}(v) \cdot  (\al_+^2(\rho_1)+\OO(p^{-1/2+\e})).
\ee
Now we can replace the estimate \eqref{bounddelta-} with this stronger estimate, and repeat all the other parts of the proof of Theorem \ref{thm_model_shift} to conclude Lemma \ref{prop_model_shift_tight}. In particular, one can calculate $  \delta^\star_{\bias}(v)$ using Lemma \ref{lem_cov_derivative} and get the $  \Delta^\star_{\bias}(v)$ term, We omit the details.
\end{proof}

>>>>

The proof of Theorem \ref{thm_model_shift} is based on Lemma \ref{lem_cov_shift}, Lemma \ref{lem_cov_derivative}, and the following bound on the singular values of $Z_1$: for any fixed $\e>0$, we have
\begin{align}
\al_-(\rho_1) - \OO(p^{-1/2+e})  \preceq {n_1^{-1}}{Z_1^T Z_1}  \preceq   \al_+(\rho_1) + \OO(p^{-1/2+e}) \quad \text{w.h.p.}  \label{eq_isometric}
\end{align}
%with high probability. 
In fact, $n_1^{-1}Z_1^TZ_1$ is a standard sample covariance matrix, and it is well-known that its eigenvalues are inside the support of the Marchenko-Pastur law $[\al_-(\rho_1)-\oo(1) ,\al_+(\rho_1)+\oo(1)]$ with probability $1-\oo(1)$ \cite{No_outside}. The estimate \eqref{eq_isometric} provides tight bounds for the concentration error. The result is formally stated in Lemma \ref{SxxSyy} below.

%\begin{remark}
The main error $\delta$ of Theorem \ref{thm_model_shift} comes from approximating $n_1^{-1}Z_1^TZ_1$ by $\id$ using \eqref{eq_isometric}; see the estimate \eqref{bounddelta-} below. In order to improve this estimate and obtain an exact asymptotic result, one needs to study the singular value distribution of the following random matrix:
$$(X_1^{\top}X_1)^{-1}X_2^{\top}X_2 +  \hat{v}^2 .$$
In fact, the eigenvalues of $\cal X:=(X_1^{\top}X_1)^{-1}X_2^{\top}X_2$ have been studied in the name of Fisher matrices; see e.g. \cite{Fmatrix}. However, since $\cal X$ is not symmetric, it is known that the singular values of $\cal X$ are different from its eigenvalues. To the best of our knowledge, the asymptotic singular value behavior of $\cal X$ is still unknown in random matrix theory literature, and the study of the singular values of $\cal X +\hat v^2$ will be even harder. We leave this problem to future study.
%\end{remark}

@@

For the isotropic model in Section \ref{sec_similarity}, we actually have an easier and sharper bound than Theorem \ref{thm_model_shift} as follows. 
\begin{lemma}[The isotropic model]\label{prop_model_shift_tight}
		In the setting of Theorem \ref{thm_model_shift}, assume that $\Sigma_1 =\id$,
		$\beta_2$ is a random vector with i.i.d. entries with mean $0$, variance $\kappa^2$ and all moments, and $\beta_1$ is a random vector such that $(\beta_1 - \beta_2)$ is a random vector with i.i.d. entries with mean $0$, variance $d^2$ and all moments. Denote
		$\Delta^\star_{\bias} := \bigbrace{(1 - \hat{v})^2 \kappa^2 + d^2} \bigtr{\Pi(\hat v)}.$
	Then we have with high probability,
		\begin{align*}
			\te(\hat{\beta}_t^{\MTL}) \le \te(\hat{\beta}_t^{\STL})\quad \text{ when: }\ \  & \Delta_{\vari} \ge  (\al_+^2(\rho_1)+\oo(1))\cdot  \Delta^\star_{\bias}, \\
			\te(\hat{\beta}_t^{\MTL}) \ge \te(\hat{\beta}_t^{\STL})\quad \text{ when: }\ \  & \Delta_{\vari} \le (\al_-^2(\rho_1)-\oo(1))\cdot  \Delta^\star_{\bias}.
		\end{align*}
\end{lemma}

\begin{proof}%[Proof of Lemma \ref{prop_model_shift_tight}]
The proof of Lemma \ref{prop_model_shift_tight} is similar to that for Theorem \ref{thm_model_shift}, except that we can replace \eqref{bounddelta-} with a tighter bound. We only describe the main difference.

For any fixed $v\in \R$, $\beta_1 - v\beta_2$ is a random vector with i.i.d. entries with mean $0$ and variance $(1-v)^2\kappa^2 + d^2$. Then using the concentration result, Lemma \ref{largedeviation}, we get that for any constant $\e>0$,
\begin{align}
&\left|\delta_{\bias}(v)- [(1-v)^2\kappa^2 + d^2]\tr (\cal K^\top \cal K)\right| \nonumber\\
&=  \left|(\beta_1 - {v} \beta_2)^\top \cal K^\top \cal K (\beta_1 - {v} \beta_2) - [(1-v)^2\kappa^2 + d^2]\tr (\cal K^\top \cal K)\right| \nonumber\\
&\le p^\e [(1-v)^2\kappa^2 + d^2] \left\{\tr\left[(\cal K^\top \cal K)^2\right]\right\}^{1/2} \lesssim p^{1/2+\e} [(1-v)^2\kappa^2 + d^2],\label{anotherbeta}
\end{align}
where we denoted $\cal K:=v\Sigma_2^{1/2}({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1} X_1^{\top}X_1$, and in the last step we used $\|\cal K\|=\OO(1)$ by \eqref{eq_isometric}. Now for $\tr (\cal K^\top \cal K)$, we rewrite it as
\begin{align*}
v^2 \tr \left[ ({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\Sigma_2({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}(X_1^\top X_1)^2 \right].
\end{align*}
Bounding $(X_1^\top X_1)^2=(Z_1^\top Z_1)^2$ using \eqref{eq_isometric} again, we obtain that
\be\label{anotherwtbeta} \delta^\star_{\bias}(v) \cdot (\al_-^2(\rho_1)-\OO(p^{-1/2+\e})) \le  [(1-v)^2\kappa^2 + d^2]\tr (\cal K^\top \cal K) \le \delta^\star_{\bias}(v) \cdot  (\al_+^2(\rho_1)+\OO(p^{-1/2+\e})),\ee
where 
$$\delta^\star_{\bias}(v):=n_1^2 v^2 [(1-v)^2\kappa^2 + d^2]\tr \left[ ({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\Sigma_2({v}^2 X_1^{\top}X_1 + X_2^{\top}X_2)^{-1}\right].$$
Note that $\delta^\star_{\bias}(v) \sim p$, hence combining \eqref{anotherbeta} and \eqref{anotherwtbeta} we get 
\be\label{replaceest}
  \delta^\star_{\bias}(v) \cdot (\al_-^2(\rho_1)-\OO(p^{-1/2+\e})) \le  \delta_{\bias}(v)\le \delta^\star_{\bias}(v) \cdot  (\al_+^2(\rho_1)+\OO(p^{-1/2+\e})).
\ee
Now we can replace the estimate \eqref{bounddelta-} with this stronger estimate, and repeat all the other parts of the proof of Theorem \ref{thm_model_shift} to conclude Lemma \ref{prop_model_shift_tight}. In particular, one can calculate $  \delta^\star_{\bias}(v)$ using Lemma \ref{lem_cov_derivative} and get the $  \Delta^\star_{\bias}(v)$ term, We omit the details.
\end{proof}
